\section{A closed-form solution for the stalling period} \label{sec:closed_form_solution}

% Based on the discussion of stalling in~\cite{DYKSTRASTALLING}, we identify as a necessary condition for stalling of the iterates for successive iteration cycles over $n$ sets. This is formalised in Definition~\ref{def:stalling}.
Based on the discussion in~\cite{DYKSTRASTALLING}, we define stalling as follows.

\begin{definition}[Stalling]\label{def:stalling}
Given $m\geq n-1$, a stalling period is defined as those $i\in\lbrace 0,\dots,m_\text{stall}\rbrace$ with $m_\text{stall}\geq n$ for which $x_{m+i}=x_{m+i-n}$.
\end{definition}

During stalling, the primary variables $x_{m+i}$ remain constant, and only the auxiliary variables $e_{m+i}$ change: these are modified by constant increments $x_m-x_{m+1}$. The stalling period therefore continues until the auxiliary variable $e_{m+i}$ becomes such that $x_{m+i}+e_{m-n+i}\in\mathcal{H}_{[m+i]}$ for some $[m]$, i.e. the half-space associated with stalling becomes inactive. For polyhedral sets, the length of the stalling period can be pre-computed as soon as the algorithm encounters stalling, using the half-spaces that are active at this iteration. The stalling period continues until one of these half-spaces becomes inactive, at which point stalling terminates. This is formalised in Theorem~\ref{thm:nstall}.

\begin{theorem}[Length of Stalling Period]\label{thm:nstall}
Suppose that stalling starts at iteration $m$ as in Definition~\ref{def:stalling} and denote the active half-spaces by $\mathcal{A}\coloneqq\set{i\in\lbrace 0,\dots,n-1\rbrace}{x_{m+i}+e_{m-n+i}\not\in\mathcal{H}_{[m+i]}}$. The length of the stalling period is
\begin{align}\label{eq:nstall}
    N_\text{stall}\coloneqq\min_{i\in\mathcal{S}} \left\lceil \frac{k_{m-n+i}}{b_{[m+i]}-x_{m+i}^T a_{[m+i]}}\right\rceil,
\end{align}
where $\lceil\cdot\rceil$ is the ceiling operator, $k_{m-n+i}=e_{m-n+i}^T a_{[m+i]}$, and $\mathcal{S}\coloneqq\set{i\in \mathcal{A}}{x_{m-n+i}^T a_{[m+i]} - b_{[m+i]} < 0}$.
\end{theorem}

\begin{proof}
Note that the existence of a finite number $N_\text{stall}$ is a consequence of the Boyle-Dykstra theorem. Following the condition in~\eqref{eq:dykstra:proj:poly}, the stalling period will terminate once an active half-space becomes inactive, which we can express as $x_{m+i}+e_{m-n+i}\in\mathcal{H}_{[m+i]}$ for some half-space index $i\geq n$ (Definition~\ref{def:stalling}). As a consequence of the scalar update in~\eqref{eq:km}, the only half-spaces that can become inactive are those for which the quantity $x_{m+i}^T a_{[m+i]} - b_{[m+i]}$ is negative. By Definition~\ref{def:stalling}, this quantity is constant during stalling. The length of the stalling period can therefore be obtained from choosing the smallest possible integer $N_i$ for which
\begin{align*}
k_{m-n+i}+N_i\left(x_{m+i}^T a_{[m+i]} - b_{[m+i]}\right) < 0,\qquad i\in\mathcal{A},
\end{align*}
so that $N_i=\left\lceil k_{m-n+i} / \left(b_{[m+i]}-x_{m+i}^T a_{[m+i]}\right)\right\rceil$, $N_\text{stall}=\min_{i\in\mathcal{S}}N_i$, and $i_\text{stall}=\Argmin_{i\in\mathcal{S}} N_i$. If multiple $i_1<\dots<i_j$ are such that $N_{i_1}=\dots = N_{i_j}$, then according to~\eqref{eq:dykstra}, half-space at index $i_\text{stall}:= i_1$ is discarded first.
\end{proof}
Note that $N_\text{stall}$ in Theorem~\ref{thm:nstall} refers to the number of $n$-step cycles of~\eqref{eq:dykstra} to skip. Following the end of the stalling period, one half-space is discarded from the active half-space and $\mathcal{A}$ redefined, upon which the algorithm may again encounter a stalling condition. By Theorem~\ref{thm:nstall}, the stalling period can be fast-forwarded by applying constant increments to the auxiliary variables $k_{m}$. Theorem~\ref{thm:nstall} and Dykstra's algorithm are combined in Algorithm~\ref{alg:fastforward}.

\algnewcommand\Inputs{\item[\textbf{Inputs:}]}
\algnewcommand\Outputs{\item[\textbf{Output:}]}

\begin{algorithm}
\caption{Dykstra's projection algorithm for polyhedral sets with fast forwarding.}
\label{alg:fastforward}
\begin{algorithmic}[1]
    \Inputs $x^\circ, \{a_i, b_i\}_{i=0}^{n-1}, N_{\max}, \epsilon_{\text{stall}}$
    \Outputs projection $x_m$
        \State $x_0 \leftarrow x^\circ$; $e_j \leftarrow 0\,\forall j\in\lbrace -n,\dots,-1\rbrace$;  $m \leftarrow 0$ 
        % \Statex\hrulefill
        \While{$m < N_{\max}$}
            \State $x_{m+1} \leftarrow \mathcal{P}_{[m]}(x_m + e_{[m-n]})$
            \State $e_{m} \leftarrow e_{m-n} + x_m - x_{m+1}$
            \If{$m \ge n-1 \;\textbf{and}\;\twonorm{x_{m+1-i} - x_{m+1-i-n}} < \epsilon_{\text{stall}}\;\forall i\in\{0,\dots,n-1\}$}\label{alg:start} \algcomment{Stalling detected per Definition~\ref{def:stalling}}
                \State Obtain $N_\text{stall}$ and $i_\text{stall}$ from Theorem~\ref{thm:nstall}
                \For{$j \in \lbrace 0,\dots,n-1\rbrace$}
                    \State $N_\text{skip}\leftarrow N_\text{stall}$ \textbf{if} $[m-j]\leq i_\text{stall}$ \textbf{else} $N_\text{stall}-1$
                    \State $e_{m-j} \leftarrow e_{m-j} + N_\text{skip}(a_{[m-j]}^T x_{m-j} - b_{[m-j]})a_{[m-j]}$\algcomment{Fast-forward auxiliary variable}
                \EndFor
                \State $m \leftarrow m - i_\text{stall}$\label{alg:end}\algcomment{Adjust iteration counter}
            \Else
                \State $m \leftarrow m + 1$
            \EndIf
        \EndWhile
        \State \Return $x_{m-1}$
\end{algorithmic}
\end{algorithm}

The algorithm proceeds identically to the standard Dykstra update until it detects stalling by verifying the stationarity condition $x_m = x_{m-n}$ for all half-spaces, as per Definition~\ref{def:stalling}. Once stalling is detected at iteration $m$, the algorithm executes its fast-forwarding phase: after performing the standard variable update for the current iteration, it computes the exact number of required stalling cycles, $N_\text{stall}$, and the half-space at which stalling ends, $i_\text{stall}$, by calling a function that implements Theorem~\ref{thm:nstall}, determining the number of full $n$-step cycles to skip. For each of the $n$ auxiliary variables $e_j$, the algorithm then computes the constant increment $\Delta e_j = (a_j^T x_{\text{stall}} - b_j)a_j$ and applies a single update: $e_j \leftarrow e_j + N_{\text{skip}} \Delta e_j$, where $N_{\text{skip}}=N_{\text{stall}}$ if $i_{\text{stall}}\geq j$ and $N_{\text{skip}}=N_{\text{stall}}-1$ otherwise. This ``jumps'' the dual variables forward to their state following the stalling period. Finally, the main iteration counter $m$ is adjusted. For practical purposes, in this implementation Algorithm~\ref{alg:fastforward} is executed for a fixed number of iterations $N_{\text{max}}$, the stationarity condition is enforced with a small nonzero tolerance $\epsilon_{\text{stall}}$, and technical details such as updating the set of active half-spaces \textit{a priori} and throughout the algorithm are omitted.

Since the proposed algorithm achieves the same iterates after escaping the stalling period (lines \ref{alg:start}--\ref{alg:end}) as would be attained with Dykstra's method, the Boyle-Dykstra theorem~\cite{DYKSTRA} is retained and the iterates are guaranteed to converge to the optimal solution $x^\star$ asymptotically. This is formalised in Corollary~\ref{cor:convergence} below.

\begin{corollary}[Convergence of Algorithm~\ref{alg:fastforward}~\cite{DYKSTRA}]
\label{cor:convergence}
The sequence of iterates $\{x_m\}$ generated by Algorithm~\ref{alg:fastforward} converges asymptotically to the optimal solution $x^\star = \mathcal{P}_{\mathcal{H}}(x^\circ)$, i.e.\ $\lim_{m\rightarrow\infty}\|x_m-x^\star\|_2=0$.
\end{corollary}